Denotational Semantics of General Payment Primitives, and Its Payment System

### 2022-10-31 packages/spec-haskell/yellowpaper/Paper1.tex
remove acks.
errata and gramatic fixes after reviews

\begin{abstract}
Payment systems in information age are still largely modeled after their analog predecessors. While
electronic money payment systems do utilize computing technology and Internet, this paper presents a
case that a true modernization can be reached by (a) making payments happening continuously over
time, (b) involving more than two parties in a payment if necessary, (c) having compositional
financial contracts.

This paper first explores the foundation of modern payment system, which consists of a money
distribution model, payment primitives, payment execution systems of financial contracts and
different forms of money mediums. Then the paper uses \textit{denotational semantics} to formally
define payment primitives for modern payment system. Lastly, this paper also includes an overview
of \textit{Superfluid protocol}, a reference implementation of the payment primitives and its
payment system.

This paper includes literate Haskell code that compiles as the constructive artifact for the formal
specifications.

This paper is also the first one in the series of yellowpapers about modern payment system dubbed
``semantic money''.

@@

@@

It should be fair to say, every aspect of money is controversial: the nature of money, the value of
money, money and banking, and monetary reconstruction. Two major schools of thoughts about theory of
money are the \textit{Austrian school} (\cite{von2013theory}) and the \textit{Chicago school}
(\cite{friedman1989quantity}). That is before the appearance of Internet-era version of monetary
reconstruction, broadly defined as cryptocurrency, which challenges theories of money further and
demands their updates (\cite{ammous2018can} \cite{hardle2020understanding}).

This yellow paper does not intend to address these controversies, instead it focuses on the function
of money. According to Von Mises:

@@

How do different forms of money perform this function, especially in the information age, when
electronic forms of money are increasingly used?

This paper adds a new controversy to money, that is to present a foundation of what constitutes of
a \textit{modern payment system} and a set of \textit{payment primitives} for the system, in order
to challenge the preconceived notions of how money can perform its function of medium of exchange.

In part \ref{part:foundation}, we shall first explore the foundation. Here we present a formal
definition of payment system and its components. We then select a few relevant approaches used in
computer science useful for modeling and defining formal specification for the payment system.

One of the approaches is \textit{denotational semantics}, which is used in part \ref{part:gpp} of
the paper to define the \textit{general payment primitives}. Along with the denotational semantics,
a restatement\footnote{It is a borrowed term from common law: ``restatement of the law''. In our
case, the denotative mathematical laws.} of it in \textit{Haskell programming language}
(\cite{hudak1992report} \cite{jones2003haskell} \cite{marlow2010haskell}) is also included.

In part \ref{part:sf}, a reference implementation of the general payment primitives and its payment
system called \textit{Superfluid Protocol} is introduced with an overview of it.

Notes on possible further investigations are also included in the end of the paper.

@@

Here we present a definition of payment system and its components.

@@

    \item \textit{payment primitives} updates money distribution,

@@

\begin{itemize}
    \item Money can be distributed continuously over time, as opposed to be in discrete chunks.

    \item Payment primitives can involve more than two parties, as opposed to be only for a sender
and a receiver.

    \item Financial system should be compositional.
\end{itemize}

@@

A representation of money and its distribution proposed in ``a unifying theory''
from \citeauthor{buldas2021unifying} involves the following components:

@@

    \item $\nu : U \rightarrow \mathbb{N}$ is the value function defining the value $\nu(u)$ of
every value unit u. The set $\mathbb{N}$ is the set of all natural numbers, but instead, we can use
any set of numerals that is totally ordered (\eg\ integers, real numbers).

    \item $\beta : U \rightarrow \mathcal{B}$ is the bearer function defining the bearer $\beta(u)$
of a unit. The set $\mathcal{B}$ is the set of possible bearers. The bearer is usually a legal
construction defining any type of legal entity, such as a person, a family, a company, a state
institution, etc.
\end{itemize}

@@

But the discrete nature of the model does not provide the necessary element for us to add the
desired properties to the payment system. Here we propose a modification that involves the usage
of \textit{context ($\gamma$)}:

@@

Note that in this model, context can be updated independently while value functions of the same
money distribution can produce different monetary values.

@@

Here are some possible components of context, and how they work:

\begin{itemize}
    \item If time (t) is included the context, then monetary value of each monetary unit can vary
continuously over time. Time is part of the physical reality, hence not changeable by the actors in
the payment system.

    \item Any subset of monetary units can also have their value functions depending on the same
information in context. This could enable payment primitives that involve many parties. This set of
information in context is referred to as \textit{ctx :: SharedContext}, they can be changed over
time by the actors in the payment system.
\end{itemize}

@@

Before the first time in this paper listing specification in Haskell, here are some highlights of
conventions in the style of the code used.

@@

In order to make the specification free of specific choices of core data types in
implementation, \textit{FunctionalDependencies}, \textit{TypeFamilies}
and \textit{TypeFamilyDependencies} are extensively used. As a consequence, the type signature can
look very cluttered. In order to address that, type quality operator $(\sim)$ is also used. Here is
an example snippets:

@@

A payment primitive is a data type with a generator function that produces payment primitive from
the shared context $ctx$, primitive specific argument $args$ and timestamp $t$, along with an update
of the shared context. Type signature \ref{eq:genPrim} is for the generator of payment
primitive \textit{a}:

@@

Loosely speaking, it is considered a primitive, if it can not be broken down into other existing
primitives which result in the same money distribution; additionally primitives should be the only
constructs in a payment system that can update money distribution.

Updates are \textit{monoidal}, so that they can be incremental and their parallel executions can be
modeled.

The most known primitive is instant transfer of monetary value between one monetary unit to
another. The introduction of context enables more primitives to be defined, and this will be
discussed in part \ref{part:gpp}.

@@

The purpose of a payment execution environment is to perform the actual payment primitives, where
their computation interface, parallel evaluation strategies and payment system solvency are defined.

It is out of scope for this paper to survey in-depth the problem space of the operational semantics
of payment execution environments. Nonetheless, a simplified model and some potential extensions to
it is discussed briefly to place payment primitives in the big picture.

@@

While money distribution does not assign meaning to the range of monetary values, in real world
applications, negative values can have special meanings. In the following analysis, we call any
money unit that has a negative monetary value \textit{insolvent}.

The detailed analysis of these solvency models are out of scope for this paper.

@@

In a nondeterministic execution environment, we cannot determine when any financial contract will
actually be executed. That means, there is always a chance a monetary unit could reach negative
monetary value.

To mitigate this uncertainty, a concept called \textit{buffer} is introduced. A buffer has a
monetary value which is set aside in a solvency conditional financial contract, such that if a
solvency condition arises, the buffer maybe drawn to cover the loss introduced by the
nondeterministic timing of the execution.

@@

Since \textit{fcNext} returns what is the next financial contract executable at a specific time,
this deterministic property eliminates the need for the buffer.

On the other hand, it introduces a different type of systemic risk, that is a kind of
denial-of-service. It is due to that the complexity of $fcUpdate$ is $O(log(n))$, the system may not
be able to advance its system time until all next executable contracts are executed.

@@

\begin{itemize}
    \item \textit{Token and its Accounts} - \eg\ bank currency accounts. Each token is its own
centralized execution environment, bearers access their monetary value through their accounts, and
execute financial contracts through the token.

    \item \textit{Note} - \eg\ federal reserve notes, bills, coins and Bitcoin UTXOs, etc. The
execution environment is independent of the notes, but it needs notes to complete the execution of
financial contracts.
\end{itemize}

One of the main differences is from the ``user interface'' perspective. A bearer expects to keep
many notes at hand, while maybe only needs a few accounts for each token. Also it is up to bearers
to keep track of all their notes, while token can keep track of most of the states for bearers;
hence notes are more decentralized and tokens are more centralized. Some also argue note-like model
is better for complex concurrent and distributed computing environment
\footfullcite[][P2]{chakravarty2020extended}.

@@

The focus of this paper is to formally define a set of payment primitives that modernize our payment
systems. To prevent reinventing wheels, it is relevant to discuss first some approaches in computer
science that are believed to help tackling this challenge.

@@

Recall that we want our modern payment system to handle money distribution continuously over time,
and its financial contracts compositional. A very closely related software design paradigm best
known to address those needs is \textit{functional reactive programming (FRP)}.  It was first
introduced by Conal Elliott \& Paul Hudak in solving multimedia animations
(\cite{elliott1997functional}). Later on, Hudak also worked on \cite{hudak2002arrows} and
\cite{wan2000functional} making FRP a more general framework for programming hybrid system with
continuous behaviors in a high-level, declarative manner.

After FRP got more adoption, it evolved to some variations that support discrete semantics, and some
variations better suited for interactive systems. In this paper we will stick to and revisit the
basic constructs of the original formulation used in \cite{elliott1997functional} where we will draw
inspirations from.

@@

Values that vary over continuous time are called \textit{behaviors}. They are first-class values,
and are built up compositionally. That is what we want in modern payment system also. The semantic
function of $\alpha$-behaviors produces the value of type $\alpha$ of a behavior at a given time:

@@

In modern payment system, the payment primitives executed in payment execution environments are
one type of events. More types of events can be read from the original paper \footfullcite[][section
2.3 Semantics of Events]{elliott1997functional}.

@@

They key to modeling the payment execution environment using FRP is the \textit{reactivity}, which
makes behaviors reactive. Specifically, the behavior \textit{b untilB e} exhibits b's behavior
until \textit{e} occurs, and then swiches to a new behavior encoded in \textit{e}:

@@

In the context of modern payment system, the occurrences of payment primitives (events) change the
behaviors of monetary units in how much monetary values it has over continuous. The building blocks
of the financial contracts should be about modeling these events and their reactivity declaratively.

@@

The syntactic category we are interested in is \textit{payment primitives}, which we should treat
them as FRP-style \textit{Behavior} data types. In the following chapters, it is also unambiguously
referred to in short as $[\![.]\!]$. Various $[\![.]\!]$ must be compositional, \ie\ must be defined
by structural recursion.

It is important to note that our purpose of using the denotational semantics is to give precise
meaning to payment primitives, independent of its implementation (which deals with performance,
optimization, side effects, etc.). We will spend the part \ref{part:gpp} exploring the denotational
semantics for general payment primitives in modern payment system.

@@

The set of payment primitives supported in a payment system is general, if (a) monetary value of
each monetary unit can vary continuously over time (b) monetary values can logically be shifted
between two or more monetary units.

The extent of the generality of each payment system may vary. This paper introduces a set of payment
primitives that is general and serves as a starting point for the readers to explore the space of
modern payment system.

@@

Recall the definition of payment system previously, money distribution sits in the core of the
system, and that is the syntactic category $\mathcal{M}$ we will be dealing with:

@@

That is to say, at any given time, the sum of monetary value of all monetary units shall always
equal to zero. Zero is used to keep the semantical meaning of payment system simple and elegant. In
its applications, a payment system may use some special monetary unit accounting negative monetary
values to accommodate concepts such as mining, minting, money printing, etc.

@@

To come up with such lawful money distribution, we can divide and conquer. In a payment system, if
we restrict that only payment primitives can provide ``updates'' to its money distribution, so that
a money distribution is only a result of a sequence of payment primitive updates; then we can have
these axioms as the basis for the proof.

@@

First we must impose the law of conservation of value on payment primitives as an axiom, in order to
prove inductively that such restricted money distribution satisfies the law of conservation of value
too.

@@

\subsection{One to One Payment Primitives}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Here are the primitives involving a sender ($u_a$) and a receiver ($u_b$) (one to one payments).

@@

A \textit{index} (we use symbol $k$ for them) has a function $\rho$ which produces a real number for
each monetary unit:

@@

Let's generalize the one to one payment primitives using index abstraction. We use $I$ subscript for
the indexed versions of the primitives.

@@

An even more general abstraction is to model participants involved a payment primitive
a \textit{network} (we use symbol $w$ for them).

It has the function $\rho$ as in \textit{index}, bnd it satisfies a different law:

@@

The question now is, should we use the most general form of the semantics to guide implementation?

The answer is most likely a no. Not because it is not useful, after all mathematical formula is
about truth and perhaps also elegance in it, but because of we may miss optimization opportunities
needed in implementation when more specialized versions are used instead\footnote{We will not though
discuss the subjective aspects, \eg\ in software engineering principles such as YAGNI
(\url{https://en.wikipedia.org/wiki/You_aren't_gonna_need_it}), or user experience perspective of
the payment primitives.}.

@@

To better serve as a reference implementation of the \textit{modern payment system} formalized by
this paper and its sequels, an implementation in Haskell was created. It aims to implement the full
specifications of (a) denotational semantics of payment primitives, (b) compositional financial
contracts for the payment primitives, (c) token \& note model of money mediums, and their execution
environment.

This paper covers the overview of the implementation with regards to (a).

@@

However, one major disadvantage is that it is not equipped with sufficient apparatus for program
correctness proofs. The best it can offer without using more experimental Haskell features is to
use \textit{QuickCheck}\footnote{\url{https://wiki.haskell.org/QuickCheck} - Haskell wiki page for
QuickCheck.} to test the necessary laws are not violated using randomized test vector.

@@

It is based on the insight of the deep connection (equivalence/isomorphism) between logic and
dependently typed programming, often called "the Curry–Howard correspondence", as discovered and
developed during the 20th century (explained in \cite{wadler2015propositions} by Phillip Wadler as
``Propositions as Types''). Agda being a dependently typed functional programming language fully
embodies this insight, hence a good candidate for creating provable correct program.

@@

\begin{displayquote}
Accounting, also known as accountancy, is the measurement, processing, and communication of
financial and non financial information about economic entities such as businesses and
corporations. \footfullcite{needles2013principles}
\end{displayquote}

The \textit{real time balance} introduced in this paper can be seen as a instance in
the \textit{general accounting domain} specifically for financial contracts in dealing with payment
primitives, which can be seen as a real time version of the cash flow accounting. Along with balance
sheet accounting, income statement accounting, the conversion between these instances of general
accounting domains are often called ``reconciliations''.

More work can be done on how to create a simple and elegant automation system for the
reconciliations.

@@

\emph{
We define the term \textbf{real time finance} to mean a financial system where its payment system is
modernized to handle money distribution continuously over time, its financial contracts are
compositional, and its general accounting domain are automated and processed in real time.  }

@@

This paper defines the what constitutes a modern payment system, with its payment primitives
formally specified using denotational semantics and restated using Haskell programming
language. Along with a reference implementation of the specification, we want to make a case that
marrying with the progress of computer science, the way of money performing its function of medium
of exchange can have a modern upgrade.

Absent of any claim on whether if such modern upgrade is needed, we invite the readers to ask
ourselves a question together: with electronic money increasingly being used, should we keep
emulating the function money of its traditional quality, or could we rethink what it may be in the
information age?

In future work, we will look into deeper other topics of real time finance, and continue to refine
the methodology used to achieve \textit{simple and precise} specifications guiding correct and
efficient engineering.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\part*{Acknowledgments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

(To be confirmed.)

>>>>

\begin{abstract}
Payment systems in the information age are still modeled after their analog predecessors. While
electronic money payment systems do utilize computing technology and the Internet, this paper
presents a case that true modernization can be reached by (a) making payments happening continuously
over time, (b) involving more than two parties in payment if necessary, (c) having compositional
financial contracts.

This paper first explores the foundation of modern payment systems, which consists of a money
distribution model, payment primitives, payment execution systems of financial contracts, and
different forms of money mediums. Then the paper uses \textit{denotational semantics} to formally
define payment primitives for modern payment systems. By the end, this paper includes an overview of
the \textit{Superfluid protocol}, a reference implementation of the payment primitives, and its
payment system.

This paper is the first in the series of yellowpapers about modern payment systems dubbed ``semantic
money.''

@@

\begin{versionhistory}
  \vhEntry{0.9}{2022-10-19}{MZC}{For reviews}
  \vhEntry{1.0}{2022-10-31}{MZC}{First release}
\end{versionhistory}

@@

It should be fair to say every aspect of money is controversial: the nature of money, the value of
money, money and banking, and monetary reconstruction. Two major schools of thought about the theory
of money are the \textit{Austrian school} (\cite{von2013theory}) and the \textit{Chicago school}
(\cite{friedman1989quantity}). That was before the appearance of the Internet-era version of
monetary reconstruction, broadly defined as cryptocurrency, which challenges theories of money
further and demands their updates (\cite{ammous2018can} \cite{hardle2020understanding}).

This yellow paper does not intend to address these controversies; instead, it focuses on the
function of money. According to Von Mises:

@@

How do different forms of money perform this function, especially in the information age, when we
increasingly use electronic forms of money?

This paper adds a new controversy to money, presenting a foundation of what constitutes
a \textit{modern payment system} and a set of \textit{payment primitives} for the system to
challenge the preconceived notions of how money can perform its function of medium of exchange.

In part \ref{part:foundation}, we shall first explore the foundation. Here we present a formal
definition of a payment system and its components. We then select a few relevant approaches used in
computer science useful for modeling and defining formal specifications for the payment system.

One of the approaches is \textit{denotational semantics}, which is used in part \ref{part:gpp} of
the paper to define the \textit{general payment primitives}. Along with the denotational semantics,
the paper also includes a restatement\footnote{It is a borrowed term from common law: ``restatement
of the law''. In our case, the denotative mathematical laws.} of it in \textit{Haskell programming
language} (\cite{hudak1992report} \cite{jones2003haskell} \cite{marlow2010haskell}).

In part \ref{part:sf}, we introduce a reference implementation of the general payment primitives and
its payment system called \textit{Superfluid Protocol}, along with its overview.

The end of the paper also includes notes on possible further investigations.

@@

Here we present a definition of a payment system and its components.

@@

    \item \textit{payment primitives} update money distribution,

@@

\begin{itemize}
    \item Money can be distributed continuously over time, as opposed to being in discrete chunks.

    \item Payment primitives can involve more than two parties, as opposed to being only for a
sender and a receiver.

    \item Financial systems should be compositional.
\end{itemize}

@@

A representation of money and its distribution proposed in ``a unifying theory''
by \citeauthor{buldas2021unifying} involves the following components:

@@

    \item $\nu: U \rightarrow \mathbb{N}$ is the value function defining the value $\nu(u)$ of every
value unit u. The set $\mathbb{N}$ is the set of all natural numbers, but instead, we can use any
set of numerals that are totally ordered (\eg,\ integers, real numbers).

    \item $\beta: U \rightarrow \mathcal{B}$ is the bearer function defining the bearer $\beta(u)$
of a unit. $\mathcal{B}$ is the set of possible bearers. The bearer is usually a legal construction
defining any type of legal entity, such as a person, a family, a company, a state institution, etc.
\end{itemize}

@@

But the discrete nature of the model does not provide the necessary element for us to add the
desired properties to the payment system. So here we propose a modification that involves the usage
of \textit{context ($\gamma$)}:

@@

Note that in this model, context can be updated independently, while value functions of the same
money distribution can produce different monetary values.

@@

Here are some possible components of context and how they work:

\begin{itemize}
    \item If time (t) is included in the context, then the monetary value of each monetary unit can
vary continuously over time. Time is part of the physical reality, hence not changeable by the
actors in the payment system.

    \item Any subset of monetary units can also have their value functions depending on the same
information in context. This could enable payment primitives that involve many parties. This set of
information in context is referred to as \textit{ctx :: SharedContext}; they can be changed over
time by the actors in the payment system.
\end{itemize}

@@

Before the first time this paper includes specification in Haskell, here are some highlights of
conventions in the code style.

@@

To make the specification free of specific choices of core data types in
implementation, \textit{FunctionalDependencies}, \textit{TypeFamilies},
and \textit{TypeFamilyDependencies} are extensively used. As a consequence, the type signature can
look very cluttered. In order to address that, type quality operator $(\sim)$ is also used. Here is
an example snippets:

@@

A payment primitive is a data type with a generator function that produces payment primitive from
the shared context $ctx$, primitive specific argument $args$, and timestamp $t$, along with an update
of the shared context. Type signature \ref{eq:genPrim} is for the generator of payment
primitive \textit{a}:

@@

Loosely speaking, it is considered primitive, if it can not be broken down into other existing
primitives, which result in the same money distribution; additionally, primitives should be the only
constructs in a payment system that can update money distribution.

Updates are \textit{monoidal} so that they can be incremental, and their parallel executions can be
modeled.

The best-known primitive is an instant transfer of monetary value between one monetary unit to
another. The introduction of context enables more primitives to be defined, and this will be
discussed in part \ref{part:gpp}.

@@

The purpose of a payment execution environment is to perform the actual payment primitives, where
their computation interface, parallel evaluation strategies, and payment system solvency are defined.

It is out of scope for this paper to survey in-depth the problem space of the operational semantics
of payment execution environments. Nonetheless, a simplified model and some potential extensions are
discussed briefly to place payment primitives in the big picture.

@@

While money distribution does not assign meaning to the range of monetary values, negative values
can have special meanings in real-world applications. In the following analysis, we call any money
unit with a negative monetary value \textit{insolvent}.

The detailed analysis of these solvency models is out of the scope of this paper.

@@

In a non-deterministic execution environment, we cannot determine when any financial contract will
be executed. That means there is always a chance a monetary unit could reach negative monetary
value.

To mitigate the uncertainty of execution time, we introduce a concept called \textit{buffer}. A
buffer has a monetary value set aside in a solvency conditional financial contract, such that if a
solvency condition arises, the buffer may be drawn to cover the loss introduced by the
non-deterministic timing of the execution.

@@

Since \textit{fcNext} returns what is the following financial contract executable at a specific
time, this deterministic property eliminates the need for the buffer.

On the other hand, it introduces a different type of systemic risk: a kind of
denial-of-service. Because the complexity of $fcUpdate$ is $O(log(n))$, the system may not be able
to advance its system time until all the following executable contracts are executed.

@@

\begin{itemize}
    \item \textit{Token and its Accounts} - \eg,\ bank currency accounts. Each token is its own
centralized execution environment; bearers access their monetary value through their accounts and
execute financial contracts through the token.

    \item \textit{Note} - \eg,\ federal reserve notes, bills, coins and Bitcoin UTXOs, etc. The
execution environment is independent of the notes, but it needs notes to complete the execution of
financial contracts.
\end{itemize}

One of the main differences is from the ``user interface'' perspective. A bearer expects to keep
many notes in hands while maybe only needing a few accounts for each token. Also, it is up to bearers
to keep track of all their notes, while tokens can keep track of most of the states for bearers;
hence notes are more decentralized and tokens are more centralized. Some also argue note-like model
is better for complex concurrent and distributed computing environment
\footfullcite[][P2]{chakravarty2020extended}.

@@

The focus of this paper is to formally define a set of payment primitives that modernize our payment
systems. However, to prevent reinventing wheels, we should first discuss some approaches in computer
science that help tackle this challenge.

@@

Recall that we want our modern payment system to handle money distribution continuously over time
and its financial contracts to be compositional. A very closely related software design paradigm
best known to address those needs is \textit{functional reactive programming (FRP)}.  It was first
introduced by Conal Elliott \& Paul Hudak in solving multimedia animations
(\cite{elliott1997functional}). Later, Hudak also worked on \cite{hudak2002arrows} and
\cite{wan2000functional}, making FRP a more general framework for programming hybrid systems with
continuous behaviors in a high-level, declarative manner.

After FRP got more adoption, it evolved into some variations that support discrete semantics and
some variations better suited for interactive systems. However, in this paper, we will stick to and
revisit the basic constructs of the original formulation used in \cite{elliott1997functional} from
which we will draw inspiration.

@@

Values that vary over continuous time are called \textit{behaviors}. They are first-class values and
are built up compositionally. That is what we want in modern payment systems also. The semantic
function of $\alpha$-behaviors produces the value of type $\alpha$ of a behavior at a given time:

@@

In modern payment systems, the payment primitives executed in payment execution environments are
one type of event. More types of events can be read from the original paper \footfullcite[][section
2.3 Semantics of Events]{elliott1997functional}.

@@

They key to modeling the payment execution environment using FRP is the \textit{reactivity}, which
makes behaviors reactive. Specifically, the behavior \textit{b untilB e} exhibits b's behavior
until \textit{e} occurs, and then switches to a new behavior encoded in \textit{e}:

@@

In the context of modern payment systems, the occurrences of payment primitives (events) change the
behaviors of monetary units in terms of how much monetary value it has over continuous. The
building blocks of the financial contracts should be about modeling these events and their
reactivity declaratively.

@@

The syntactic category we are interested in is \textit{payment primitives}, which we should treat as
FRP-style \textit{Behavior} data types. In the following chapters, it is also unambiguously referred
to in short as $[\![.]\!]$. Various $[\![.]\!]$ must be compositional, \ie,\ must be defined by
structural recursion.

It is important to note that our purpose in using the denotational semantics is to give precise
meaning to payment primitives independent of their implementations (which deal with performance,
optimization, side effects, etc.). We will spend part \ref{part:gpp} exploring the denotational
semantics for general payment primitives in modern payment systems.

@@

The set of payment primitives supported in a payment system is general if (a) the monetary value of
each monetary unit can vary continuously over time (b) monetary values can logically be shifted
between two or more monetary units.

The extent of the generality of each payment system may vary. This paper introduces a set of payment
primitives that is general and serves as a starting point for the readers to explore the space of
modern payment systems.

@@

Recall in the definition of payment system previously, money distribution sits in the core of the
system, and that is the syntactic category $\mathcal{M}$ we will be dealing with:

@@

That is to say, at any given time, the sum of the monetary value of all monetary units shall always
equal zero. Zero is used to keep the semantical meaning of payment systems simple and elegant. In
its applications, a payment system may use some special monetary unit accounting for negative
monetary values to accommodate concepts such as mining, minting, money printing, etc.

@@

To come up with such lawful money distribution, we can divide and conquer. In a payment system, if
we restrict that only payment primitives can provide ``updates'' to their money distribution so that
money distribution is only a result of a sequence of payment primitive updates; then, we can have
these axioms as the basis for the proof.

@@

First, we must impose the law of conservation of value on payment primitives as an axiom in order to
prove inductively that such restricted money distribution satisfies the law of conservation of value
too.

@@

\subsection{One-to-One Payment Primitives}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Here are the primitives involving a sender ($u_a$) and a receiver ($u_b$) (one-to-one payments).

@@

An \textit{index} (we use symbol $k$ for them) has a function $\rho$ which produces a real number for
each monetary unit:

@@

Let's generalize the one-to-one payment primitives using index abstraction. We use $I$ subscript for
the indexed versions of the primitives.

@@

An even more general abstraction is to model participants involved a payment
primitive \textit{network} (we use the symbol $w$ for them).

It has the function $\rho$ as in \textit{index}, and it satisfies a different law:

@@

The question now is, should we use the most general form of semantics to guide implementation?

The answer is most likely a no. Not because it is not useful; after all mathematical formula is
about truth and perhaps also elegance in it, but because of we may miss optimization opportunities
needed in implementation when more specialized versions are used instead\footnote{We will not though
discuss the subjective aspects, \eg\ in software engineering principles such as YAGNI
(\url{https://en.wikipedia.org/wiki/You_aren't_gonna_need_it}), or user experience perspective of
the payment primitives.}.

@@

To better serve as a reference implementation of the \textit{modern payment system} formalized by
this paper and its sequels, an implementation in Haskell was created. It aims to implement the full
specifications of (a) the denotational semantics of payment primitives, (b) compositional financial
contracts for the payment primitives, (c) token \& note model of money mediums, and their execution
environment.

This paper covers the overview of the implementation with regard to (a).

@@

However, one major disadvantage is that it is not equipped with sufficient apparatus for program
correctness proofs. The best it can offer without using more experimental Haskell features is to
use \textit{QuickCheck}\footnote{\url{https://wiki.haskell.org/QuickCheck} - Haskell wiki page for
QuickCheck.} to test that the necessary laws are not violated using randomized test vectors.

@@

It is based on the insight of the deep connection (equivalence/isomorphism) between logic and
dependently typed programming, often called "the Curry–Howard correspondence", as discovered and
developed during the 20th century (explained in \cite{wadler2015propositions} by Phillip Wadler as
``Propositions as Types''). Agda being a dependently typed functional programming language fully
embodies this insight, hence a good candidate for creating provable correct programs.

@@

\begin{displayquote}
Accounting, also known as accountancy, is the measurement, processing, and communication of
financial and non-financial information about economic entities such as businesses and
corporations. \footfullcite{needles2013principles}
\end{displayquote}

The \textit{real time balance} introduced in this paper can be seen as an instance in
the \textit{general accounting domain}, specifically for financial contracts in dealing with payment
primitives, which can be seen as a real time version of cash flow accounting. Along with balance
sheet accounting, and income statement accounting, the conversion between these instances of general
accounting domains are often called ``reconciliations''.

More work can be done on how to create a simple and elegant automation system for reconciliations.

@@

\emph{
We define the term \textbf{real time finance} to mean a financial system where its payment system is
modernized to handle money distribution continuously over time, its financial contracts are
compositional, and its general accounting domain is automated and processed in real time.  }

@@

This paper defines what constitutes a modern payment system, with its payment primitives formally
specified using denotational semantics and restated using Haskell programming language. Along with a
reference implementation of the specification, we want to make a case that marrying with the
progress of computer science, the way of money performs its function of medium of exchange can have
a modern upgrade.

Absent any claim on whether a modern upgrade is needed; we invite the readers to ask ourselves a
question together: with electronic money increasingly being used, should we keep emulating the
function of money of its traditional quality, or could we rethink what it may be in the information
age?

In future work, we will look into deeper other topics of real time finance and continue to refine
the methodology used to achieve \textit{simple and precise} specifications guiding correct and
efficient engineering.
